Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-10 analysis/paper/paper.Rmd
edit first half of discussion section

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans (Cheng, 2008; Wang, 2011), hinting at an emergence of social inequality in the Indigenous community. We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation (Costin, 2001; Junker, 1999). The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay (cf. Kwak and Marwick, 2015). Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation (cf. Evershed, 2008, p. 909).

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population (Chen, 2007; Hsieh, 2009).  We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure 7) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

>>>>

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at an emergence of social inequality in the Indigenous community. We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, (Figure \@ref(fig:kernel-plot)) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
